### 2019-11-15 paper/paper.tex
Friday work on the paper

We first verified that the system given by Eqns. \ref{eq:dc} \& \ref{eq:da}
described a two dimensional extension of the axon branching model in
\citet{karbowski_model_2004}. We showed that in the presence of three guidance
gradients varying only in one dimension (along the long axis of an elliptical
domain), five thalamocortical axon types, each interacting uniquely with the
three guidance gradients, would segregate into five separate regions. We
investigated a number of configurations of the guidance gradients, varying
their shape, $\rho(\mb{x})$, and number, $M$, and their effect on a varying
number of thalamocortical projections, $N$. In general, we were not able to
producen a complexity of structure in the resulting patterns of the
connections, $c(\mb{x},t\rightarrow\infty)$, which was not present in the
guidance patterns.

Complex structure can indeed be observed in protein expression for some genes,
such as the molecules Bhlhb5 \citep{joshi_bhlhb5_2008} and Lmo4
\citep{cederquist_lmo4_2013,greig_molecular_2013}. These expression patterns
can therefore provide a theoretical mechanism for the guidance of axons into
distinct regions.
%The question of how such structures are generated becomes
%one of of how gene transcriptional interactions and mRNA diffusion and
%movement achieves this (\cmnt{Fixme: find a ref on structure in gene
%  transcription}).
However, complex and stable structure in gene expression data is not commonly
reported; monotonically varying expression patterns, such as those for Pax6,
Emx2, Sp8 and Coup-tf1, which display graded expression across the entire
isocortex (\cmnt{REFS}) are more typical. Furthermore, the complex structure
observed in the whisker barrel field has not, to our knowledge, been linked to
a set of gene expression patterns matching the scale of the barrels. To
investigate how the whisker barrel structure might form in the presence of
only monotonically varying morphogen gradients, we turned our attention to the
competition for connections built into the model.

@@

There exists an element of competition in the model system specified by
Eqns. \ref{eq:dc} \& \ref{eq:da}. The term $\left(1 - \sum_{i=1}^{N}
c_i(\mb{x}, t)\right)$ in Eq.~\ref{eq:dc} ensures that once the available
synaptic receptors in a region have been fully occupied, say by
thalamocortical type $i=1$, there is no longer the possibility for type $i=2$
to form any synapses. This competition turns out to be relatively weak,
because it does not strongly affect the branching density in that region. In
the model, it is possible for two or more TC axon types to have roughly equal
branching density in some region, leading to roughly equal numbers of
connections for each TC type being made, up until the `hard' connection limit
is reached.

To introduce competition between the \emph{branching} of thalamocortical axon
types, we introduced the new term $\chi$ given by
%
\begin{equation} \label{eq:branch_comp}
\chi(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{i' \ne i}^{N} a_{i'}(\mb{x}, t)
\end{equation}
%
which leads to a reduction in the rate of increase of branching density for
axon type $i$ in regions where the branching density of \e{other} axon types
is high.

@@

We first investigate the effect of the competition by examining a system of 25
thalamocortical axon types simulated on a square domain. The axons interact
with two orgthogonal signalling gradients, one of which increases from the top
edge to the bottom edge, the other of which increases from the right edge to
the left edge. The values of $\gamma$ are chosen as a grid.

>>>>

We first verified that the system given by Eqs.~\ref{eq:dc} \& \ref{eq:da}
described a two dimensional extension of the axon branching model in
\citet{karbowski_model_2004}. We showed that in the presence of three guidance
gradients varying only in one dimension (along the long axis of an elliptical
domain), five thalamocortical axon types, each interacting uniquely with the
three guidance gradients, would segregate into five separate regions. We
investigated a number of configurations of the guidance gradients, varying
their shape, $\rho(\mb{x})$, and number, $M$, and their effect on a varying
number of thalamocortical projections, $N$. In general, we were not able to
producen a complexity of structure in the resulting patterns of the
connections, $c(\mb{x},t\rightarrow\infty)$, which was not present in the
guidance patterns.

Complex structure can indeed be observed in protein expression for some genes,
such as the molecules Bhlhb5 \citep{joshi_bhlhb5_2008} and Lmo4
\citep{cederquist_lmo4_2013,greig_molecular_2013}. These expression patterns
can therefore provide a theoretical mechanism for the guidance of axons into
distinct regions.
%The question of how such structures are generated becomes
%one of of how gene transcriptional interactions and mRNA diffusion and
%movement achieves this (\cmnt{Fixme: find a ref on structure in gene
%  transcription}).
However, complex and stable structure in gene expression data is not commonly
reported; monotonically varying expression patterns, such as those for Pax6,
Emx2 \citep{shimogori_fibroblast_2005,bishop_regulation_2000},
Sp8~\citep{sahara_sp8_2007} and Coup-tf1~\citep{borello_sp8_2014}, which
display graded expression across the entire isocortex are more
typical. Furthermore, the complex structure observed in the whisker barrel
field has not, to our knowledge, been linked to a set of gene expression
patterns matching the scale of the barrels. To investigate how the whisker
barrel structure might form in the presence of only monotonically varying
morphogen gradients, we turned our attention to the competition for
connections built into the model.

@@

There exists an element of competition in the model system specified by
Eqs.~\ref{eq:dc} \& \ref{eq:da}. The term $\left(1 - \sum_{i=1}^{N}
c_i(\mb{x}, t)\right)$ in Eq.~\ref{eq:dc} ensures that once the available
synaptic receptors in a region have been fully occupied, say by
thalamocortical type $i=1$, there is no longer the possibility for type $i=2$
to form any synapses. This competition turns out to be relatively weak,
because it does not strongly affect the branching density in that region. In
the model, it is possible for two or more TC axon types to have roughly equal
branching density in some region, leading to roughly equal numbers of
connections forming for each TC type, up until the `hard' connection limit
is reached.

To introduce competition between the \emph{branching} of thalamocortical axon
types, we introduced the new term $\chi$ given by
%
\begin{equation} \label{eq:branch_comp}
\chi(\mb{x}, t) = - \frac{\epsilon  a_i(\mb{x}, t)}{N-1} \sum_{i' \ne i}^{N} a_{i'}(\mb{x}, t).
\end{equation}
%
which leads to a reduction in the rate of increase of branching density for
axon type $i$ in regions where the branching density of \e{other} axon types
is high.

@@

With the addition of the competition term, $\chi$, it is necessary to
introduce a normalisation step, in order to prevent one TC type from
out-competing every other TC type. Normalisation conforms to the idea that for
a fixed number of afferent thalamocortical axons, the overall number of
branches must have some upper limit; given the presumably finite resources
available in each axon. We divisively normalize the branching density,
dividing $a_i$ by the mean of $a_i$ per unit area; formalised as:
%
\begin{equation} \label{eq:norm}
a_i(\mathbf{x}, t) = A \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega a'_i(\mathbf{x}, t) d\sigma}
\end{equation}
%
where the prime symbols indicate the use of the intermediate value of $a_i$
after it has been computed according to Eqs.~\ref{eq:dc}--\ref{eq:branch_comp}
on the basis of the system state at time $t-{\delta}t$. $A$ is the area of the
domain. This normalises $a_i(\mathbf{x}, t)$ to 1, integrated over the domain. To
account for the possibility that there might be more axons afferent for TC
type $1$ cf.~TC type $2$, we normalise to the integral of $a_i$ at time 0.
%
\begin{equation} \label{eq:norm_time0}
\begin{split}
a_i(\mathbf{x}, t) & =
\frac{\int_\Omega  a'_i(\mathbf{x}, 0) d\sigma}{A} \; A \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma} \\
%
& = \int_\Omega  a'_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma}
\end{split}
\end{equation}
%
In the simulation, in which space is represented as $n_e$ discrete
hexagonal tiles, the equivalent expression is:
%
\begin{equation} \label{eq:norm_discrete}
{a_{i,p}}(t) = \textstyle \sum_j^{n_e} a_{i,j}(0) \; \frac {a'_{i,p}(t)} {\textstyle \sum_j^{n_e} a'_{i,j}(t) }
\end{equation}
%
where $a_{i,j}(t)$ is the value of the variable $a_i$ in the $j^{\mathrm{th}}$
hexagonal element at time t.

We first investigated the effect of the competition by examining a system of
25 thalamocortical axon types simulated on a square domain, as in
Fig.~\ref{fig:square}. The hex to hex distance was 0.02 mm and the
square's side length was 1.16 mm. The axons interacted with two orthogonal
signalling gradients, one of which increased from the top edge to the bottom
edge, the other of which had its gradient increasing from the right edge to
the left edge [Fig.~\ref{fig:square}(f)]. For each thalamocortical type, $i$,
two interaction parameters were set; $\gamma_{i,1}$ was the interaction with
guidance molecule 1 and $\gamma_{i,2}$ was the interaction with molecule
2. The branching of TC types with a negative interaction with molecule 2
($\gamma_{i,2}<0$) will be pushed to the left; those with $\gamma_{i,2}>0$ are
pushed to the right. TC types that interact negatively with molecule 1
($\gamma_{i,1}<0$) are pushed up; those with $\gamma_{i,1}>0$ are pushed down.

Fig.~\ref{fig:square}(a) shows the connection densities after the first 100
simulation time steps (t=0.005 s). Each colour uniquely identifies the TC
type, $i$ [the full mapping between index $i$ and the colour is shown in
  Fig.~\ref{fig:square}(d)]. The left-hand map colours each hex according to
which TC type has the highest connection density. At this early stage of the
simulation, the colouring is somewhat random because the branching density is
initialised stochastically ($a_{i,j}(0) = 0.2 + U([0,0.2])$), which results in
a randomised pattern for ${c_i}^{max}$. The right-hand map of
Fig.~\ref{fig:square}(a) colours those hexes which have $c_i \geq 0.5$ and an
adjoining hex for which $c_i < 0.5$ according to the mapping shown in
Fig.~\ref{fig:square}(d)]. Due to the initial stochasticity, there are many
  such crossings of the threshold. The predominance of blue (for $i<10$)
  results from the order in which the hexes have been rendered.
Fig.~\ref{fig:square}(b) shows the simulation at t=0.14. \cmnt{CONTINUE FROM HERE.}

% Later
The full set of these interaction parameters are shown in
Fig.~\ref{fig:square}(e). The values of $\gamma$ were chosen with the
intention of forming a grid; the interactions for $i =$ 1, 5, 21 \& 25 (the
corners) were strongest and the interaction parameters for $i=$13 (the centre)
were zero. Fig.~\ref{fig:square}(d) shows the mapping between colour and TC
type $i$ and can be used, in conjunction with Fig.~\ref{fig:square}(e), to see
that $\gamma_{1,1}=-2$, $\gamma_{1,2}=-2$, $\gamma_{3,1}=0$,
$\gamma_{3,2}=-2$, and so on.
